---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  dpi = 300,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/",
  warning = FALSE,
  message = FALSE 
)
```

# nzffdr <img src="man/figures/nzffdr_hex.svg" align="right" height="150"/></a>

<!-- badges: start -->

[![CRAN status](https://www.r-pkg.org/badges/version/nzffdr)](https://CRAN.R-project.org/package=nzffdr) [![R-CMD-check](https://github.com/flee598/nzffdr/workflows/R-CMD-check/badge.svg)](https://github.com/flee598/nzffdr/actions) [![CRAN RStudio mirror downloads](https://cranlogs.r-pkg.org/badges/grand-total/nzffdr)](https://r-pkg.org/pkg/nzffdr)

<!-- badges: end -->

The purpose of this package is to allow for direct access to the NZ Freshwater Fish Database ([NZFFD](https://nzffdms.niwa.co.nz/search)) from R and additional functions for cleaning imported data and adding missing data.

Specifically, the seven core functions are:

-   `nzffdr_import()` - import NZFFD data into R.

-   `nzffdr_clean()` - clean up a bunch of small inconsistencies in the imported NZFFD data.

-   `nzffdr_add_dates()` - add separate year, month and day columns to NZFFD data.

-   `nzffdr_taxon_threat()` - add extra taxonomic and NZ Threat Classification System ([NZTCS](https://nztcs.org.nz)) data to NZFFD data (12 new columns). Lots more detail added compared to `nzffdr` v 1.0.0.

-   `nzffdr_widen_habitat()` - transform non-tidy format habitat columns to tidy wide format.

-   `nzffdr_razzle_dazzle()` - wrapper for all of the above functions to quickly download and tidy NZFFD data in one go.

-   `nzffdr_ind_lengths()` - transform non-tidy format individual fish length measures into tidy long format.

See below for a detailed overview.

### Installation

Install from CRAN
```{r, warning = FALSE, eval = FALSE}
install.packages("nzffdr")
```

Or to install the latest development version:

```{r, warning = FALSE, eval = FALSE}
devtools::install_github("flee598/nzffdr")
```

### Built-in datasets

There are two built-in datasets, these are:

-   `nzffdr_data` a subset of 200 rows from the NZFFD, used for examples, tutorials etc.

-   `nzffdr_nzmap` a simple features map of New Zealand. A simplified version of the 1:150k NZ map outline available from [Land Information New Zealand](https://www.linz.govt.nz).

### Getting data

As of Novemeber 2022 there are issues with the NIWA portal (see notes in changelog), it is best to download the entire NZFFD database (e.g. use `import()` with all args empty, although you can set `download_format`) and then filter as needed in R.

We have tried to make the search terms match those you would use directly on the NIWA site. For example leaving a search field blank will return all records. `nzffdr_import()` has nine search arguments:

-   `institution` the organisation that submitted the records.

-   `catchment_num` this refers to the Catchment No., a 6 digit number unique to the reach of interest. You can search using the individual number (e.g. `catchment = "702.500"`), or for all rivers in a catchment you can use the wildcard search term (e.g. `catchment = "702%"`), or don't set the arg if you want all catchments in NZ.

-   `catchment_name` search for a catchment by name e.g. (`catchment_name = "Waimakariri R"`).

-   `waterBody` search for a water body by name. (e.g. `waterBody = "Otukaikino Stream tributary"`).

-   `fish_method` search by fishing method used. If you want to search for a specific fishing method use the function `nzffdr_get_table("fish_method")` to see a list of all possible search terms, (e.g. if you only wanted fish caught by lures use `fish_meth = "Angling - Lure"`).

-   `taxon` search for a particular species. If you want to search for a specific species use the function `nzffdr_get_table("taxon")` to see a list of all possible search terms. You can search using either common or scientific names and can search for multiple species at once. e.g. to search for Black mudfish use `species = "Black mudfish"` or `species = "Neochanna diversus"` and to search for Black mudfish and Bluegill bully use `species = c("Black mudfish", "Bluegill bully")` etc.

-   `starts` start search date. Don't set the arg if you want all records in the database.

-   `ends` end search date. Don't set the arg if you want all records in the database.

-   `download_format` can be either `all` or `essential` depending on how many columns of data you want.

This function requires an internet connection to query NIWA's database.

Data citation: Stoffels R (2022). New Zealand Freshwater Fish Database (extended). The National Institute of Water and Atmospheric Research (NIWA). Sampling event dataset <https://doi.org/10.15468/jbpw92>

```{r}
library(nzffdr)

# import all records - all fields
dat <- nzffdr_import(download_format = "all")

# To import the entire NZFF database:
# dat <- nzffdr_import()
```

### Cleaning data

While the data imported from NZFFD is in pretty good shape there are some small inconsistencies. The `nzffd_clean()` function aims to deal with some of these inconsistencies.Column types are checked and converted to, integer, numeric or character. Empty cells are filled with `NA`, variable `catchmentName` entries are standardised (e.g. `Clutha r`, `Clutha River` and `Clutha R` all become `Clutha R`) and, any empty columns are removed.

```{r}
dat2 <- nzffdr_clean(dat)
```

The above changes, while superficial make analysis that, for example relies on grouping, work as intended.

### Add separate date columns

A small function that adds separate year, month and day columns based on values in the column `eventDate`.

```{r}
dat3 <- nzffdr_add_dates(dat2)

# Have a look at the new columns
dat3[1:5, c("year", "month", "day")]
```

### Add taxonomic and threat status information

Add NZ Threat Classification System ([NZTCS](https://nztcs.org.nz)) information for each of the species. Includes taxonomic data, common and some MÄori names. Also threat status (e.g. declining) and whether the species is endemic, invasive etc.

```{r}
dat4 <- nzffdr_taxon_threat(dat3)

# Have a look at the new columns
dat4[1:5, c("commonMaoriName", "alternativeNames",
            "genus", "family", "order", "class", "phylum",
            "category", "status", "taxonomicStatus", "bioStatus")]
```

### Tidy habitat columns

The NZFFD now comes with additional habitat information in the columns (`habitatFlowPercent`, `habitatInstreamCoverPresent`, `habitatRiparianVegPercent`, `habitatSubstratePercent`), by default there are multiple entries in a single cell. `nzffdr_widen_habitat()` transforms these non-tidy format habitat columns to tidy wide format columns.

```{r}
dat5 <- nzffdr_widen_habitat(dat4)

# Have a look at the new columns
dat5[1:5, c("Fine_gravel", "Coarse_gravel", "Sand_1_2_mm", "Mud", 
            "Cobbles_64_257_mm", "Boulders_257_mm", "Bedrock", "Gravel_3_64_mm",
            "Mud_Silt_1mm", "Run", "Riffle", "Pool", "Still", "Backwater",
            "Rapid", "Cascade", "Torrent", "Grass_tussock", "Scrub",
            "Native_forest", "Raupo_flax", "Exotic_forest", "Exposed_bed",
            "Other", "Tussock", "Scrub_willow", "Pasture", "Undercut_banks",
            "Wood_instream_debris", "Bank_vegetation", "Macrophytes_algae",
            "Macrophytes", "Cobbles", "Periphyton")]
```

### Quick all in one

There is a wrapper for all of the above functions, to quickly download the entire NZFFD, tidy and add information all in one easy step.

```{r}
dat_all <- nzffdr_razzle_dazzle()

# total number of rows and columns in the dataset
dim(dat_all)

# have a look at the first 5 rows and 10 columns
dat_all[1:5, 1:10]

```

### Mapping observations

There is a simple features map of New Zealand included in the package, this can be used to quickly check species distributions and the like.

```{r}
# load ggplot2 and sf for nice figures
library(ggplot2)
library(sf)

# get map of NZ and remove Chatham Islands
nz <- nzffdr::nzffdr_nzmap
nz <- subset(nz, name != "Chatham Island")

# filter just Oncorhynchus genus 
df_fish <- subset(dat_all, genus == "Oncorhynchus")

# create a basic map, with points coloured by species common names
ggplot() +
  geom_sf(data = nz,
          fill = "white") +
  geom_point(data = df_fish, aes(x = eastingNZTM, y = northingNZTM, colour = taxonCommonName),
             alpha = 0.6) +
  scale_colour_brewer(palette = "Dark2") +
  labs(colour = expression(italic("Oncorhynchus"))) +
  theme_void() +
  theme(plot.background = element_rect(fill = 'lightgrey', colour = 'lightgrey'),
        legend.position=c(.3, .8),
        text = element_text(size = 8))
```
