---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  dpi = 600,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  warning = FALSE,
  message = FALSE 
)
```

# nzffdr <img src='man/figures/nzffdr_hex.svg' align="right" height="150" /></a>


<!-- badges: start -->
[![R-CMD-check](https://github.com/flee598/nzffdr/workflows/R-CMD-check/badge.svg)](https://github.com/flee598/nzffdr/actions)
<!-- badges: end -->


## Major update December 2021
Following a significant update to the NZFFD, the nzffdr package as been completely overhauled. The NZFFD update resulted in nzffdr v 1.0.0 functions either i) not working or, ii) being redundant. Therefore I have completely rewritten nzffdr. This does mean there is *NO* backwards compatibility, and any previous code that used nzffdr functions will no longer work. I am sorry for this, but it was unavoidable. 

On the up side, the NZFFD now contains lots of useful additional information by default, and the new and improved nzffdr has added functionality.

## Package overview

The purpose of this package is to allow for direct access to the NZ Freshwater Fish Database ([NZFFD](https://nzffdms.niwa.co.nz/search)) from R and additional functions for cleaning imported data and adding missing data. 

For a detailed guide to using the package see the [help page](https://flee598.github.io/nzffdr/).

A preprint describing the package is now available (if you wish to cite the package please use this):

Lee, F., & Young, N. (2021). nzffdr: an R package to import, clean and update data from the New Zealand Freshwater Fish Database. bioRxiv. https://doi.org/10.1101/2021.06.22.449519

### Installation
For now, the only functioning version of nzffdr is here on GitHub, I aim to get nzffdr v 2.0.0 on CRAN in early 2022.

### Development version
To install the latest development version install from GitHub.
```{r, warning = FALSE, eval = FALSE}
devtools::install_github("flee598/nzffdr")
```

### Import data from the NZFFD 

This function requires an internet connection to query the NZFFD database. 

```{r}
# import all records between 2010 and 2015
library(nzffdr)

dat <- nzffdr_import(institution = "", catchment_num = "", catchment_name = "",
                    water_body = "", fish_method = "", taxon = "", 
                    starts = 2010, ends = 2015, download_format = "all")
head(dat)

# To import the entire NZFF Database:
# dat <- nzffd_import()
```