---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  dpi = 600,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  warning = FALSE,
  message = FALSE 
)
```

# nzffdr <img src='man/figures/nzffdr_hex.svg' align="right" height="150" /></a>


<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/nzffdr)](https://CRAN.R-project.org/package=nzffdr)
[![R-CMD-check](https://github.com/flee598/nzffdr/workflows/R-CMD-check/badge.svg)](https://github.com/flee598/nzffdr/actions)
  [![CRAN RStudio mirror downloads](https://cranlogs.r-pkg.org/badges/grand-total/nzffdr)](https://r-pkg.org/pkg/nzffdr)
<!-- badges: end -->


## Package overview

The purpose of this package is to allow for direct access to the NZ Freshwater Fish Database ([NZFFD](https://nzffdms.niwa.co.nz/search)) from R and additional functions for cleaning imported data and adding missing data. 

For a detailed guide to using the package see the [help page](https://flee598.github.io/nzffdr/).

A preprint describing the package is now available (if you wish to cite the package please use this):

Lee, F., & Young, N. (2021). nzffdr: an R package to import, clean and update data from the New Zealand Freshwater Fish Database. bioRxiv. https://doi.org/10.1101/2021.06.22.449519

### Installation
```{r, warning = FALSE, eval = FALSE}
install.packages("nzffdr")
```


### Development version
To install the latest development version install from GitHub.
```{r, warning = FALSE, eval = FALSE}
devtools::install_github("flee598/nzffdr")
```

### Import data from the NZFFD 

This function requires an internet connection to query the NZFFD database. 

```{r}
# import all records 
library(nzffdr)

dat <- nzffdr_import(download_format = "essential")
head(dat)

# To import the entire NZFF Database:
# dat <- nzffd_import()
```