---
output: github_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  dpi = 600,
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  warning = FALSE,
  message = FALSE 
)
```

# nzffdr <img src='man/figures/nzffdr_hex.svg' align="right" height="150" /></a>


<!-- badges: start -->
[![CRAN status](https://www.r-pkg.org/badges/version/nzffdr)](https://CRAN.R-project.org/package=nzffdr)
[![R-CMD-check](https://github.com/flee598/nzffdr/workflows/R-CMD-check/badge.svg)](https://github.com/flee598/nzffdr/actions)
  [![CRAN RStudio mirror downloads](https://cranlogs.r-pkg.org/badges/grand-total/nzffdr)](https://r-pkg.org/pkg/nzffdr)
<!-- badges: end -->

## Major update December 2021
Following a significant update to the NZFFD, the nzffdr package has been completely overhauled. The NZFFD update resulted in nzffdr v 1.0.0 functions either i) not working or, ii) being redundant. Therefore I have completely rewritten nzffdr. This does mean there is **NO** backwards compatibility. I am sorry for this, but it was out of my control given the changes to NZFFD. 

On the up side, the NZFFD now contains lots of useful additional information by default, and the new and improved nzffdr has added functionality.

## Package overview

The purpose of this package is to allow for direct access to the NZ Freshwater Fish Database ([NZFFD](https://nzffdms.niwa.co.nz/search)) from R and additional functions for cleaning imported data and adding missing data. 

For a detailed guide to using the package see the [help page](https://flee598.github.io/nzffdr/).

A preprint describing the package is now available (if you wish to cite the package please use this):

Lee, F., & Young, N. (2021). nzffdr: an R package to import, clean and update data from the New Zealand Freshwater Fish Database. bioRxiv. https://doi.org/10.1101/2021.06.22.449519

### Installation
nzffdr v 2.0.0 is now available on CRAN.
```{r, warning = FALSE, eval = FALSE}
install.packages("nzffdr")
```


### Development version
To install the latest development version install from GitHub.
```{r, warning = FALSE, eval = FALSE}
devtools::install_github("flee598/nzffdr")
```

### Import data from the NZFFD 

This function requires an internet connection to query the NZFFD database. 

```{r}
# import all records between 2010 and 2015
library(nzffdr)

dat <- nzffdr_import(institution = "", catchment_num = "", catchment_name = "",
                    water_body = "", fish_method = "", taxon = "", 
                    starts = 2010, ends = 2015, download_format = "all")
head(dat)

# To import the entire NZFF Database:
# dat <- nzffd_import()
```